---
title: "Biostat 203B Homework 5"
subtitle: Due Mar 22 @ 11:59PM
author: "Xinrui Xiong 806329308"
---

## Predicting ICU duration

Using the ICU cohort `mimiciv_icu_cohort.rds` you built in Homework 4, develop at least three machine learning approaches (logistic regression with enet regularization, random forest, boosting, SVM, MLP, etc) plus a model stacking approach for predicting whether a patient's ICU stay will be longer than 2 days. You should use the `los_long` variable as the outcome. You algorithms can use patient demographic information (gender, age at ICU `intime`, marital status, race), ICU admission information (first care unit), the last lab measurements before the ICU stay, and first vital measurements during ICU stay as features. You are welcome to use any feature engineering techniques you think are appropriate; but make sure to not use features that are not available at an ICU stay's `intime`. For instance, `last_careunit` cannot be used in your algorithms.

1.  Data preprocessing and feature engineering.

```{r, eval = T}
library(tidyverse)
library(tidymodels)
library(ggplot2)
library(GGally)
library(gtsummary)
library(keras)
library(ranger)
library(stacks)
library(vip)

```

```{r, eval=T}
mimiciv_icu_cohort <- readRDS("./mimic_icu_cohort.rds")

mimiciv_icu_cohort$los_long <- as.factor(mimiciv_icu_cohort$los_long)

mimiciv_icu_cohort <- mimiciv_icu_cohort %>%
  select(subject_id, hadm_id, stay_id, gender, 
         age_intime, marital_status, race, first_careunit,
         `White Blood Cells`, `Creatinine`, `Potassium`, 
         `Glucose`, `Sodium`, `Chloride`, `Bicarbonate`, 
         `Hematocrit`, `Heart Rate`, 
         `Non Invasive Blood Pressure systolic`,
         `Temperature Fahrenheit`, `Respiratory Rate`, 
         `Non Invasive Blood Pressure diastolic`, los_long)


tbl_summary(mimiciv_icu_cohort %>% 
              select(
                -subject_id, 
                -hadm_id, 
                -stay_id), 
            by=los_long)

```

From the tbl_summary, we can see that there are some missing values in the dataset. We'll use the mean value to fill in the missing values for the continuous variables and use the mode value to fill in the missing values for the categorical variables in the recipe procedure in the next step.

2.  Partition data into 50% training set and 50% test set. Stratify partitioning according to `los_long`. For grading purpose, sort the data by `subject_id`, `hadm_id`, and `stay_id` and use the seed `203` for the initial data split. Below is the sample code.

```{r, eval = T}
set.seed(203)

# sort
mimiciv_icu_cohort <- mimiciv_icu_cohort |>
  arrange(subject_id, hadm_id, stay_id)

data_split <- initial_split(
  mimiciv_icu_cohort, 
  # stratify by los_long
  strata = "los_long", 
  prop = 0.5
  )


mimic_training<-training(data_split)
mimic_testing<-testing(data_split)

```

```{r, eval = T}
# recipe
# To exclued the subject_id, hadm_id, stay_id, "los_long ~ . "are not used.
recipe <- recipe(
  los_long ~ gender + age_intime + marital_status + 
    race + first_careunit + `White Blood Cells` + 
    `Creatinine` + `Potassium` + `Glucose` + `Sodium` + 
    `Chloride` + `Bicarbonate` + `Hematocrit` + 
    `Heart Rate` + `Non Invasive Blood Pressure systolic` + 
    `Temperature Fahrenheit` + `Respiratory Rate` + 
    `Non Invasive Blood Pressure diastolic`, 
  data = mimic_training) |>
  # mean imputation for continuous variables
  step_impute_mean(`White Blood Cells`, `Creatinine`, 
                   `Potassium`, `Glucose`, `Sodium`, 
                   `Chloride`, `Bicarbonate`, `Hematocrit`, 
                   `Heart Rate`, `Non Invasive Blood Pressure systolic`, 
                   `Temperature Fahrenheit`, `Respiratory Rate`, 
                   `Non Invasive Blood Pressure diastolic`) |>
  # mode imputation for categorical variables
  step_impute_mode(marital_status) |>
  # create traditional dummy variables
  step_dummy(all_nominal_predictors()) |>
  # zero-variance filter
  step_zv(all_numeric_predictors()) |> 
  # center and scale numeric data
  step_normalize(all_numeric_predictors()) |>
  # estimate the means and standard deviations
  # prep(training = mimic_training, retain = TRUE) |>
  print()
  

```

3.  Train and tune the models using the training set.

**logistic regression with enet regularization:**

```{r, eval = T}

logit_mod <- logistic_reg(
  penalty = tune(), 
  mixture = tune()) %>%
  set_engine("glmnet") %>%
  set_mode("classification") %>%
  print()

```

```{r, eval = T}
logit_wf <- workflow() |>
  add_recipe(recipe) |>
  add_model(logit_mod) |>
  print()
```

```{r, eval = T}
param_grid <- grid_regular(
  penalty(range = c(-6, 3)), 
  mixture(),
  levels = c(100, 5)
  ) |>
  print()
```

```{r, eval = T}

set.seed(203)

folds <- vfold_cv(mimic_training, v = 5)
folds

```

```{r, eval = T}
logit_fit <- logit_wf |>
  tune_grid(
    resamples = folds,
    grid = param_grid,
    metrics = metric_set(roc_auc, accuracy)
    )

logit_fit
```

```{r, eval = T}
logit_fit |>
  # aggregate metrics from K folds
  collect_metrics() |>
  print(width = Inf) |>
  filter(.metric == "roc_auc") |>
  ggplot(mapping = aes(
    x = penalty, 
    y = mean, 
    color = factor(mixture)
    )) +
  geom_point() +
  labs(x = "Penalty", y = "CV AUC") +
  scale_x_log10()
```

```{r, eval = T}
logit_fit |>
  show_best(metric = "roc_auc")


best_logit <- logit_fit |>
  select_best(metric="roc_auc")
best_logit
```

```{r, eval = T}
# Final workflow
logit_final_wf <- logit_wf |>
  finalize_workflow(best_logit)
logit_final_wf

# Fit the whole training set, then predict the test cases
logit_final_fit <- 
  logit_final_wf |>
  last_fit(data_split)
logit_final_fit

# Test metrics
logit_final_fit |> 
  collect_metrics()
```

**Random Forest**

```{r, eval = T}
rf_mod <- 
  rand_forest(
    mode = "classification",
    # Number of predictors randomly sampled in each split
    mtry = tune(),
    # Number of trees in ensemble
    trees = tune()
  ) |> 
  set_engine("ranger")
rf_mod

```

```{r, eval = T}
rf_wf <- workflow() |>
  add_recipe(recipe) |>
  add_model(rf_mod) |>
  print()
```

```{r, eval = T}
param_grid <- grid_regular(
  trees(range = c(100L, 200L)), 
  mtry(range = c(1L, 5L)),
  levels = c(5, 5)
  )
param_grid
```

```{r, eval = T}
set.seed(203)

folds <- vfold_cv(mimic_training, v = 5)
folds


rf_fit <- rf_wf |>
  tune_grid(
    resamples = folds,
    grid = param_grid,
    metrics = metric_set(roc_auc, accuracy)
    )
rf_fit
```

```{r, eval = T}
rf_fit |>
  collect_metrics() |>
  print(width = Inf) |>
  filter(.metric == "roc_auc") |>
  ggplot(mapping = aes(
    x = trees, y = mean, color = factor(mtry)
    )) +
  geom_point() + 
  # geom_line() + 
  labs(x = "Num. of Trees", y = "CV AUC")

rf_fit |>
  show_best(metric = "roc_auc")


best_rf <- rf_fit |>
  select_best(metric = "roc_auc")
best_rf
```

```{r, eval = T}
# Final workflow
rf_final_wf <- rf_wf |>
  finalize_workflow(best_rf)
rf_final_wf


# Fit the whole training set, then predict the test cases
rf_final_fit <- 
  rf_final_wf |>
  last_fit(data_split)
rf_final_fit


# Test metrics
rf_final_fit |> 
  collect_metrics()
```

**boosting**

```{r, eval = T}
gb_mod <- 
  boost_tree(
    mode = "classification",
    trees = 100, 
    tree_depth = tune(),
    learn_rate = tune()
  ) |> 
  set_engine("xgboost")
gb_mod
```

```{r, eval = T}
gb_wf <- workflow() |>
  add_recipe(recipe) |>
  add_model(gb_mod)
gb_wf

```

```{r, eval = T}
param_grid <- grid_regular(
  tree_depth(range = c(1L, 3L)),
  learn_rate(range = c(-5, 2), 
             trans = log10_trans()),
  levels = c(3, 10)
  )
param_grid
```

```{r, eval = T}

set.seed(203)

folds <- vfold_cv(mimic_training, v = 5)
folds

gb_fit <- gb_wf |>
  tune_grid(
    resamples = folds,
    grid = param_grid,
    metrics = metric_set(roc_auc, accuracy)
    )
gb_fit

```

```{r, eval = T}
gb_fit |>
  collect_metrics() |>
  print(width = Inf) |>
  filter(.metric == "roc_auc") |>
  ggplot(mapping = aes(
    x = learn_rate, 
    y = mean, 
    color = factor(tree_depth)
    )) +
  geom_point() +
  labs(x = "Learning Rate", y = "CV AUC") +
  scale_x_log10()

gb_fit |>
  show_best(metric = "roc_auc")

best_gb <- gb_fit |>
  select_best(metric = "roc_auc")
best_gb
```

```{r, eval = T}
# Final workflow
gb_final_wf <- gb_wf |>
  finalize_workflow(best_gb)
gb_final_wf

# Fit the whole training set, then predict the test cases
gb_final_fit <- 
  gb_final_wf |>
  last_fit(data_split)
gb_final_fit

# Test metrics
gb_final_fit |> 
  collect_metrics()
```

**Model Stacking (Ensemble Learning)**

```{r, eval = T}
logit_grid <- grid_regular(
  penalty(range = c(-6, 3)), 
  mixture(),
  levels = c(100, 5)
  )

logit_res <- 
  tune_grid(
    object = logit_wf, 
    resamples = folds, 
    grid = logit_grid,
    control = control_stack_grid()
  )
logit_res


rf_grid <- grid_regular(
  trees(range = c(100L, 500L)), 
  mtry(range = c(1L, 5L)),
  levels = c(5, 5)
  )

rf_res <- 
  tune_grid(
    object = rf_wf, 
    resamples = folds, 
    grid = rf_grid,
    control = control_stack_grid()
  )
rf_res

gb_grid <- grid_regular(
  tree_depth(range = c(1L, 3L)),
  learn_rate(range = c(-5, 2), trans = log10_trans()),
  levels = c(3, 10)
  )

gb_res <- 
  tune_grid(
    object = gb_wf, 
    resamples = folds, 
    grid = gb_grid,
    control = control_stack_grid()
  )
```

```{r, eval = T}
mimic_model_st <- 
  # initialize the stack
  stacks() |>
  # add candidate members
  add_candidates(logit_res) |>
  add_candidates(rf_res) |>
  add_candidates(gb_res) |>
  # determine how to combine their predictions
  blend_predictions(
    penalty = 10^(-6:2),
    metrics = c("roc_auc")
    ) |>
  # fit the candidates with nonzero stacking coefficients
  fit_members()
mimic_model_st

autoplot(mimic_model_st)
autoplot(mimic_model_st, type = "members")
autoplot(mimic_model_st, type = "weights")


collect_parameters(mimic_model_st, "logit_res")
collect_parameters(mimic_model_st, "rf_res")
collect_parameters(mimic_model_st, "gb_res")


```

```{r, eval = T}
mimic_pred <- mimic_testing %>%
  bind_cols(predict(mimic_model_st, ., type = "prob")) %>%
  print(width = Inf)


mimic_pred <- mimic_pred %>%
  mutate(.pred_class = if_else(.pred_FALSE > 0.5, 
                               "FALSE", 
                               "TRUE"))%>%
  mutate(.pred_class = factor(.pred_class,
                              levels = c("FALSE", "TRUE")))


mimic_pred$los_long <- factor(mimic_pred$los_long, 
                              levels = c("FALSE", "TRUE"))
  
```

```{r, eval = T}
yardstick::roc_auc(
  mimic_pred,
  truth = los_long,
  contains(".pred_FALSE")
  )

yardstick::accuracy(
  mimic_pred,
  truth = los_long,
  estimate = .pred_class
  )
```

4.  Compare model classification performance on the test set. Report both the area under ROC curve and accuracy for each machine learning algorithm and the model stacking. Interpret the results. What are the most important features in predicting long ICU stays? How do the models compare in terms of performance and interpretability?

<font color ="Blue"> **Firstly, the performance are concluded as follows:**

**Logistic Regression with Elastic Net Regularization:**

-   AUC: 0.5965392\
-   Accuracy: 0.5736409

**Random Forest:**

-   AUC: 0.6329192
-   Accuracy: 0.5961586

**Boosting:**

-   AUC: 0.6386602
-   Accuracy: 0.6008920

**Model Stacking:** - AUC: 0.6461363 - Accuracy: 0.6054612


To determine the most important features in predicting long ICU stays, we can use the `vip` package to calculate the variable importance. 
Let's use the boosting model as an example:

```{r, eval = T}
final_tree <- extract_workflow(gb_final_fit)
final_tree 

final_tree %>% 
  extract_fit_parsnip() %>% 
  vip()

```
As the plot shows, the Temperature Fahrenheit is the most important feature in predicting long ICU stays.



The model stacking has the best performance and interpretability: AUC is a measure of how well the model can distinguish between the two classes. The higher the AUC, the better the model is at predicting the outcome. The AUC for the model stacking is 0.6461363, which is the highest among all the models. This indicates that the model stacking is the best at distinguishing between the two classes. The accuracy of the model stacking is 0.6054612, which is also the highest among all the models. This indicates that the model stacking is the best at predicting the outcome. As for the rest 3 models, their performance can be ranked as follows: Boosting \> Random Forest \> Logistic Regression with Elastic Net Regularization.

</font>
